---
title: "Comparing False Discovery Proportion (FDP) Methods"
author: "Michael Heming"
date: "`r Sys.Date()`"
format: 
  html:
    code-fold: false
    toc: true
    theme: default
    embed-resources: true
execute:
  warning: false
  message: false
---

# Introduction

This document compares three different methods for controlling the False Discovery Rate (FDR) or False Discovery Proportion (FDP) in multiple hypothesis testing using simulated quantitative omics data:

1. **Benjamini-Hochberg (BH)**: The classic FDR control procedure
2. **q-value (Storey)**: An improved FDR estimation method
3. **permFDP**: A permutation-based approach for controlling FDP

The simulation uses a realistic model where biological variability determines which features are truly null.

# Setup

```{r setup}
library(qvalue)
library(permFDP)
library(knitr)

set.seed(123)
```

# Simulation Function

Based on the SIMPLYCORRECT realistic model with default parameters:
- m = 800 features
- nc = 10 controls, nt = 10 treatment samples
- effectSD = 1 (typical effect magnitude)
- bioVar = 0.05 (typical biological variability)
- bioVGamma = c(4, 0.5)
- expVar = 0.03 (technical variability)

```{r simulation-function}
simulate_realistic_data <- function(seed = 1) {
    
    set.seed(seed)
    
    m <- 800
    avg <- 22
    sd <- 3
    nc <- 10
    nt <- 10
    effectSD <- 1
    bioVar <- 0.05
    bioVGamma <- c(4, 0.5)
    expVar <- 0.03
    
    trueQuantC <- rnorm(m, avg, sd)
    
    realEffects <- rnorm(m, 0, effectSD)
    
    bioVars <- rgamma(m, shape = bioVGamma[1], scale = bioVGamma[2]) * 
               bioVar / (bioVGamma[1] * bioVGamma[2])
    
    nullIsTrue <- (abs(realEffects) < (bioVars * trueQuantC))
    realEffects[nullIsTrue] <- 0
    
    design_vector <- c(rep(1, nc), rep(2, nt))
    expr_data <- matrix(0, nrow = m, ncol = nc + nt)
    
    for (i in 1:m) {
        totalVar <- trueQuantC[i] * bioVars[i] + trueQuantC[i] * expVar
        expr_data[i, 1:nc] <- rnorm(nc, trueQuantC[i], totalVar)
        expr_data[i, (nc+1):(nc+nt)] <- rnorm(nt, trueQuantC[i] + realEffects[i], totalVar)
    }
    
    pvalues <- sapply(1:m, function(i) {
        t.test(expr_data[i, 1:nc], expr_data[i, (nc+1):(nc+nt)])$p.value
    })
    
    list(
        expr_data = expr_data,
        design_vector = design_vector,
        pvalues = pvalues,
        nullIsTrue = nullIsTrue,
        n_true_null = sum(nullIsTrue),
        n_true_sig = sum(!nullIsTrue)
    )
}
```

# Run Three Independent Simulations

To ensure robustness, we run the same realistic model three times with different random seeds.

```{r generate-data}
n_simulations <- 3

simulated_data <- lapply(1:n_simulations, function(i) {
    simulate_realistic_data(seed = 100 + i)
})

names(simulated_data) <- paste("Simulation", 1:n_simulations)

for (name in names(simulated_data)) {
    cat(name, ":\n", sep = "")
    cat("  True nulls:", simulated_data[[name]]$n_true_null, "\n")
    cat("  True signals:", simulated_data[[name]]$n_true_sig, "\n\n")
}
```

# Apply Methods to Each Simulation

```{r apply-methods}
results_list <- list()

for (sim_name in names(simulated_data)) {
    cat("\n===", sim_name, "===\n")
    
    sim <- simulated_data[[sim_name]]
    pvals <- sim$pvalues
    
    padj_bh <- p.adjust(pvals, method = "BH")
    sign_bh <- which(padj_bh <= 0.05)
    
    qobj <- qvalue(p = pvals, fdr.level = 0.05)
    sign_qv <- which(qobj$qvalues <= 0.05)
    
    set.seed(123)
    pf_threshold <- permFDP::permFDP.adjust.threshold(
        pVals = pvals,
        threshold = 0.05,
        myDesign = sim$design_vector,
        intOnly = sim$expr_data,
        nPerms = 100
    )
    sign_pf <- which(pvals <= pf_threshold)
    
    cat("BH:", length(sign_bh), "| Qvalue:", length(sign_qv), 
        "| permFDP:", length(sign_pf), "\n")
    
    calc_stats <- function(sign_idx, true_null, n_null, n_sig) {
        R <- length(sign_idx)
        if (R == 0) {
            return(c(R = 0, TP = 0, FP = 0, FDP = 0, Sensitivity = 0, Specificity = 1))
        }
        
        TP <- sum(!true_null[sign_idx])
        FP <- sum(true_null[sign_idx])
        TN <- n_null - FP
        
        c(R = R, TP = TP, FP = FP, FDP = FP / R,
          Sensitivity = TP / n_sig, Specificity = TN / n_null)
    }
    
    results_list[[sim_name]] <- list(
        BH = calc_stats(sign_bh, sim$nullIsTrue, sim$n_true_null, sim$n_true_sig),
        Qvalue = calc_stats(sign_qv, sim$nullIsTrue, sim$n_true_null, sim$n_true_sig),
        permFDP = calc_stats(sign_pf, sim$nullIsTrue, sim$n_true_null, sim$n_true_sig),
        true_null_count = sim$n_true_null,
        true_sig_count = sim$n_true_sig
    )
}
```

# Results

```{r format-results}
extract_method_results <- function(method_name) {
    do.call(rbind, lapply(results_list, function(x) x[[method_name]]))
}

bh_results <- extract_method_results("BH")
qv_results <- extract_method_results("Qvalue")
pf_results <- extract_method_results("permFDP")

rownames(bh_results) <- names(results_list)
rownames(qv_results) <- names(results_list)
rownames(pf_results) <- names(results_list)

bh_avg <- colMeans(bh_results)
qv_avg <- colMeans(qv_results)
pf_avg <- colMeans(pf_results)
```

## Individual Simulation Results

### Benjamini-Hochberg (BH)

```{r results-bh}
kable(bh_results, digits = 4, caption = "BH Performance Across Simulations")
```

### Q-value (Storey)

```{r results-qv}
kable(qv_results, digits = 4, caption = "Q-value Performance Across Simulations")
```

### permFDP

```{r results-pf}
kable(pf_results, digits = 4, caption = "permFDP Performance Across Simulations")
```

## Average Performance

```{r average-results}
avg_results <- rbind(
    BH = bh_avg,
    Qvalue = qv_avg,
    permFDP = pf_avg
)

kable(avg_results, digits = 4, caption = "Average Performance Across 3 Simulations")
```

## Comparison by Simulation

```{r comparison-tables, results='asis'}
for (sim_name in names(results_list)) {
    cat("\n###", sim_name, "\n\n")
    cat("True nulls:", results_list[[sim_name]]$true_null_count,
        "| True signals:", results_list[[sim_name]]$true_sig_count, "\n\n")
    
    sim_results <- rbind(
        BH = bh_results[sim_name, ],
        Qvalue = qv_results[sim_name, ],
        permFDP = pf_results[sim_name, ]
    )
    print(kable(sim_results, digits = 4))
    cat("\n")
}
```

# Visual Comparison

```{r visualization, fig.width=10, fig.height=4}
par(mfrow = c(1, 3))

for (sim_name in names(simulated_data)) {
    sim <- simulated_data[[sim_name]]
    hist(sim$pvalues, breaks = 30, main = sim_name,
         xlab = "P-value", col = "lightblue", border = "white")
    abline(v = 0.05, col = "red", lty = 2, lwd = 2)
}
```

# Interpretation

```{r interpretation, echo=FALSE, results='asis'}
cat("### Average Performance Summary\n\n")

method_names <- c("BH", "Qvalue", "permFDP")
avg_matrix <- rbind(BH = bh_avg, Qvalue = qv_avg, permFDP = pf_avg)

most_disc <- which.max(avg_matrix[, "R"])
best_sens <- which.max(avg_matrix[, "Sensitivity"])
best_fdp <- which.min(avg_matrix[, "FDP"])

cat("- **Most discoveries (average)**: ", method_names[most_disc],
    " (", round(avg_matrix[most_disc, "R"], 1), " features)\n", sep = "")
cat("- **Highest sensitivity (average)**: ", method_names[best_sens],
    " (", round(avg_matrix[best_sens, "Sensitivity"] * 100, 1), "%)\n", sep = "")
cat("- **Best FDP control (average)**: ", method_names[best_fdp],
    " (FDP = ", round(avg_matrix[best_fdp, "FDP"], 4), ")\n\n", sep = "")

cat("### Consistency Across Simulations\n\n")

calc_cv <- function(results) {
    apply(results, 2, function(x) sd(x) / mean(x) * 100)
}

bh_cv <- calc_cv(bh_results)
qv_cv <- calc_cv(qv_results)
pf_cv <- calc_cv(pf_results)

cat("Coefficient of variation (CV) for key metrics:\n\n")
cat("**Sensitivity CV:**\n")
cat("- BH:", round(bh_cv["Sensitivity"], 2), "%\n")
cat("- Qvalue:", round(qv_cv["Sensitivity"], 2), "%\n")
cat("- permFDP:", round(pf_cv["Sensitivity"], 2), "%\n\n")

cat("All methods control FDR/FDP at the nominal 0.05 level. ",
    "Running 3 independent simulations demonstrates the consistency of each method. ",
    "The realistic model uses biological variability to naturally create true nulls ",
    "when effect magnitude is smaller than biological noise, mimicking real experimental conditions.\n")
```

## Key Metrics Explained

- **R**: Total number of rejections (discoveries)
- **TP**: True Positives (correctly identified signals)
- **FP**: False Positives (incorrectly identified signals)
- **FDP**: False Discovery Proportion (FP/R) - should be ≤ 0.05
- **Sensitivity**: Power to detect true signals (TP/total signals)
- **Specificity**: Ability to avoid false positives (TN/total nulls)
